Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-05 talk/dls2014/paper/paper.tex
add a diagram for object states

The \textbf{young generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of a thread's nursery into the old object space, which can be done
without stopping other threads. This is done either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

@@

@@

The job of the write barrier is twofold: first, it serves as a write
barrier for the garbage collector and second, it supports COW and adds
objects to the write set of the transaction.

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER!  in the object's
header through its $SO$ and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier
transaction, or if there was a minor collection which will add the
flag again on all objects. It is never set on freshly allocated
objects that still reside in the nursery.

@@

The \textbf{slow path} is shown here in Listing~\ref{lst:w_slow}.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to.

The check for \lstinline!is_overflow_obj()! looks at the
\lstinline!overflow_number!  and tells us if the object was actually
created in this transaction. In that case, we do not need to execute
the following \emph{TM part}.  We especially do not need to privatise
its pages since no other transaction knows about these overflow
objects. Even if they reside in non-private pages, it is guaranteed
that no other transaction can have a reference to them.

For the \emph{TM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \textbf{global} array of bytes that is indexed with the
$SO$ of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we will do a write-write contention
management that will abort either us or the current owner of the
object.  If we succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

\cfbolz{at this point I kind of had the need for a diagram of the various
states an object can be in, and the transitions between them.}

@@

\arigo{what else than memory usage goes here?}

>>>>

The \textbf{young generation}, where objects are considered to be
\emph{young} and reside in the \emph{Nursery}, is collected by
\emph{minor collections}. These collections move the surviving objects
out of a thread's nursery into the old object space, which can be done
without stopping other threads. This is happens either if the nursery has
no space left anymore or if we are committing the current
transaction. Consequently, all objects are old and the nursery empty
after a transaction commits.  Furthermore, all objects in the nursery
were always created in the current transaction. This fact is useful
since we do not need to call any barrier on this kind of objects.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{object_states.pdf}
  \caption{Objects states handled by the write barrier\label{fig:obj_states}}
\end{figure}

@@

The job of the write barrier is twofold: first, it serves as a write
barrier for the garbage collector and second, it supports COW and adds
objects to the write set of the transaction. These two jobs are
depicted in Figure~\ref{fig:obj_states}.

The \textbf{fast path} of the write barrier is very simple. We only
need to check for the flag \lstinline!WRITE_BARRIER! in the object's
header through its $SO$ and call the slow path if it is set. This flag
is set either if the object is old and comes from an earlier
transaction, or if there was a minor collection which will add the
flag again on all objects and may produce overflow objects. The flag
is never set on freshly allocated objects, because they do not need
to be traced by the GC or handled by the STM system. This fast path
covers all cases for the GC and for the STM system.

@@

The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to.

The check for \lstinline!is_overflow_obj()! looks at the
\lstinline!overflow_number!  and tells us if the object was actually
created in this transaction. In that case, we do not need to execute
the following \emph{STM part}.  We especially do not need to privatise
its pages since no other transaction knows about these overflow
objects. Even if they reside in non-private pages, it is guaranteed
that no other transaction can have a reference to them.

Up to here, we see that all cases of the GC shown in
Figure~\ref{fig:obj_states} are handled. What is left is to handle the
cases of the STM system that need to make old objects writable.

For the \emph{STM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \textbf{global} array of bytes that is indexed with the
$SO$ of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we will do a write-write contention
management that will abort either us or the current owner of the
object.  If we succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

\arigo{what else than memory usage goes here?}
\remi{limited scalability; static segmentation?}
